---
title: "Bayesian modelling part4"#using waitt_samoa file
author: "Fimareti Selu"
date: today
date-format: "DD/MM/YYYY"
format: 
  html:
    ## Format
    theme: spacelab
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    fig-caption-location: "bottom"
    fig-align: "center"

    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    highlight-style: zenburn
    ## Execution
    execute:
      echo: true
      cache: false
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```

# Preparations
Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

library(tidyverse)
library(easystats)
library(knitr)
library(sf)
library(rnaturalearth)
library(maps)
library(brms)
library(rstan)
library(tidybayes)
library(patchwork)
library(DHARMa)
library(HDInterval)
library(emmeans)
source('helperFunctions.R')
```

    
# Read in the data

Now we will move to the raw `CairnsReefs_subset.csv` data.
There are many functions in R that can read in a CSV file. We will use
a the `read_csv()` function as it is part of the tidyverse ecosystem.

```{r}
#| label: readData
dat <- read_csv("../data/Samoa_Waitt_updated.csv", trim_ws = TRUE)
```

```{r}
#| label: readLabelset
labelset <- read_csv("../data/ReefCloudlabelset.csv", trim_ws = TRUE) 
```


::: {.panel-tabset}

## glimpse
```{r}
#| label: examinData
dat |> glimpse() 
```

## head
```{r}
#| label: examinData1
## Explore the first 6 rows of the data
dat |> head() 
```

## str
```{r}
#| label: examinData2
dat |> str() 
```

## Easystats (datawizard)
```{r}
#| label: examinData3
dat |> datawizard::data_codebook() |> knitr::kable() 
```
:::


# Data preparation

Before we can model these data, they need to be processed into a
format compatible with statistical modelling. The necessary wrangling
steps:

1. exclude extraneous (unneeded) fields
2. exclude poor images
3. lengthen the data with respect to classification type
4. join to a labelset lookup
5. tally up the points per date/image/GROUP/type
6. recode transect id
7. fill in the gaps and add the zeros
8. sum to transect level
9. generate a Year field from the sample date

::: {.panel-tabset}

## Exclude fields

Although it is often harmless enough to retain the other fields, it
does make reviewing the data more combersum, so at an early stage
within this exercise, we will probably restrict the data to just the
above fields.

```{r}
#| label: selecting 
dat <- dat |>
  dplyr::select(site_id,
    site_name,
    site_latitude,
    site_longitude,
    survey_start_date,
    survey_depth,
    survey_transect_number,
    image_id,
    image_quality,
    point_id,
    point_num,
    point_machine_classification,
    point_human_classification
    )
dat |> as.data.frame() |> head()
```


## Excluding poor images

```{r}
#| label: filter
dat <-
  dat |>
  dplyr::filter(is.na(image_quality) | image_quality != 0)  
dat |> as.data.frame() |> head()
dat |> dim()
```


## Lengthen the data

```{r}
#| label: pivot
dat <-
  dat |>
  pivot_longer(cols = matches("point_.*_classification"),
    names_to = "type",
    values_to = "classification"
    ) 
dat |> as.data.frame() |> head()
```

## Joining to the group code lookup data
Our primary interest in these data will be in exploring patterns in
broad taxanomic groups such as *Hard coral* and *algae*. Since the
machine learning tools behind ReefCloud are tuned to classify to finer
taxonomic resolutions, next task will be to use a lookup table so as
to assign the higher order group codes (HC and A) to the data.


:::: {.panel-tabset}

### glimpse
```{r}
#| label: examineLabelset
labelset |> glimpse() 
```

### head
```{r}
#| label: examineLabelset1
## Explore the first 6 rows of the labelseta
labelset |> head() 
```

### str
```{r}
#| label: examineLabelset2
labelset |> str() 
```

### Easystats (datawizard)
```{r}
#| label: examineLabelset3
labelset |> datawizard::data_codebook() |> knitr::kable() 
```
::::

```{r}
#| label: join
dat <-
  dat |>
  left_join(labelset |>
              dplyr::select(CODE, GROUP = `FUNCTIONAL GROUP`),
              by = c("classification" = "CODE")
    ) |> 
  mutate(GROUP = ifelse(str_detect(GROUP, "HC_.*"), "HC", GROUP))
dat |> as.data.frame() |> head() 
```

## Exclude rows without classification
```{r}
dat <- dat |> 
  filter(!is.na(classification)) |> 
  droplevels()
```

## Tally up points

Count the number of points of each type as well as sum up the total
number of points per image.

```{r}
#| label: count
dat <- 
  dat |> 
  group_by(across(c(starts_with("site"),
    starts_with("survey"),
    type,
    image_id,
    GROUP))
  ) |>
  summarise(COUNT = n(), .groups = "keep") |> 
  ungroup(GROUP) |>
  mutate(TOTAL = sum(COUNT)) |>
  ungroup() 
dat |> as.data.frame() |> head() 
```

## Recode transects

```{r}
#| label: recode_transects
dat <- 
  dat |>
  mutate(transect_id = paste0(site_id, survey_transect_number)) 
dat |> as.data.frame() |> head() 
```

## Fill in any gaps

Since the data represent the classification of points in images, they
only include what was present, not what was also absent. For example,
if all five points are Algae, then this also means that all other
functional groups are absent - yet this information is lacking in the
data. For modelling purposes it is vital that we fill in all the zero
values.

To do so, we must create a data set that contains every GROUP in every
IMAGE.


```{r}
GROUPS <- dat |> pull(GROUP) |> unique()
data.filler <- dat %>%
  dplyr::select(
    starts_with("site"),
    survey_start_date,
    #Year,
    survey_depth,
    transect_id,
    image_id,
    type,
    ## GROUP,
    TOTAL) |> 
  distinct() |> 
 tidyr::crossing(GROUP = GROUPS) 

dat <-
  dat |> 
  full_join(data.filler) |>
  group_by(
    across(c(starts_with("site"),
      survey_start_date,
      #Year,
      survey_depth,
      transect_id,
      image_id,
      type,
      GROUP
    ))) |> 
  mutate(COUNT = ifelse(is.na(COUNT), 0, COUNT),
    TOTAL = max(TOTAL, na.rm = TRUE)
  )
dat |> as.data.frame() |> head() 
```

## Sum to transect level


```{r}
## Now sum up to transect level
dat <- 
  dat |>
  ungroup(image_id) |>
  summarise(COUNT = sum(COUNT),
    TOTAL = sum(TOTAL)
  ) |> 
  ungroup() |> 
  droplevels()
dat |> as.data.frame() |> head() 
```

## Generate a year field

```{r}
#| label: mutateYear
dat <-
  dat |>
  mutate(Year = lubridate::year(survey_start_date),
    TropYear = lubridate::year(survey_start_date + months(3))
  ) 
dat |> as.data.frame() |> head() 
```

## Generate Reef_id

```{r}
#| label: mutataReef_id
dat <-
  dat |>
  mutate(Reef_id = str_replace(site_name, "(.*)_.*", "\\1"))

dat |> as.data.frame() |> head() 
```
## Create a variable to compare two groups of sites
```{r}
dat <- dat |> 
  mutate(SAV_UPO = str_replace(site_name, "(.*)_.*", "\\1"),
         SAV_UPO = factor(SAV_UPO))
```


## Declare all character vectors as categorical

We will also create a categorical version of year.

```{r}
#| label: declare factors
dat <-
  dat |>
  mutate(across(where(is.character), factor))

dat |> as.data.frame() |> head() 
```
##Mapping
```{r}
reefs <- read_sf("../data/GIS/reef_500_poly.shp")
```

```{r}
cities<- maps::world.cities |> 
  filter (country.etc == "Samoa") |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)
Samoa<- rnaturalearth::ne_countries(scale = 10, country = "Samoa", returnclass = "sf")
#Samoa
bbox<- Samoa |> st_bbox()
ggplot()+
  geom_sf(data=Samoa)+
  geom_sf(data=reefs, fill = "lightblue") +
  geom_sf(data = cities) +
  #geom_point(data=cities, aes(y= lat, x=long))+
  #geom_text(data = cities, aes(y = lat,x = long-0.05,label = name), hjust=1)+
  ggspatial::annotation_north_arrow(location= "tr")+
  ggspatial::annotation_scale()+
  coord_sf(xlim=bbox[c("xmin", "xmax")],ylim = bbox[c("ymin", "ymax")])+
  theme_bw()+
  theme(axis.title = element_blank())
```

:::

## Exploratory data analysis

::: {.panel-tabset}

### Time series plot

```{r}
#| label: EDA1
#| fig.width: 8
#| fig.height: 4

dat |>
  filter(type == "point_machine_classification", GROUP == "HC") |> 
  ggplot(aes(y =  100*COUNT/TOTAL, x = SAV_UPO)) +
  geom_point() +
  #geom_line(aes(group = transect_id)) + 
  scale_y_continuous("Hard coral cover (%)") +
  #scale_colour_discrete("Survey depth (m)") +
  #scale_x_datetime("Year", date_breaks = "1 years", date_labels = "%Y") + 
  facet_wrap(~site_name) +
  theme_classic()
```
```{r}
dat_hc <- dat |>
  filter(GROUP == "HC") |>
  droplevels()

```

### Boxplots

```{r}
#| label: EDA2
#| fig.width: 6
#| fig.height: 4
dat |>
  filter(type == "point_machine_classification", GROUP == "HC") |> 
  ggplot(aes(y =  100*COUNT/TOTAL, x = SAV_UPO)) +
  geom_boxplot() 
```

**Conclusions -** variances are clearly not equal between the groups
:::

Benthic monitoring programs typically adopt a hierarchical sampling
design in which (usually) fixed sampling units (such as quadrats or
transects) are nested within in Sites which are in turn nested within
progressively larger spatial units (e.g. Reefs within Regions). In
such designs, monitors collect data from those same units over
multiple years.  That is, the units are repeatedly measured.

Such a design is an attempt to control some of the numerous sources of
uncertainty that would otherwise make it very difficult to draw out a
signal (temporal pattern) amongst the noise. If instead, researchers
sampled different locations each year, it is much harder to uncouple
how much of any changes are due to genuine temporal shifts and how
much is due to just a shift in location.

Whist this design does provide substantial statistical power
advantages, it does introduce additional complexities to do with the
independence of observations. Repeated observations from the same
sampling units (transects etc) are not independent. Similarly,
observations of sampling units that are close in space (or time) are
likely to be highly correlated. This violates a important assumption
that underlies most statistical analyses.

Complex hierarchical statistical designs exist to accommodate these
dependency issues. Although such analyses are very powerful, they are
not a good starting point when learning how to fit models. This series
of workshops is intended to progressively increase the complexity of
the statistical designs and analyses.

The current dataset represents only a tiny fraction of a much larger
dataset that includes multiple sites, reefs and years. As such, it has
been prepared to fill in for a very simple design/analyses.

::: {.callout-important}
Unfortunately, these data are still not appropriate for a very simple
analyses for the following two reasons:

- as we can see in the time series figure above, the earlier samples
  were collected at a depth of 9m whereas the later samples were
  collected at 7m
- the samples each year are not independent as they were collected
  from the same transects - that is the sampling units have been
  repeatedly sampled and thus are not independent.

I fully acknowledge this issue and therefore acknowledge that the
analysis I will present is completely invalid.
:::


```{r, mhiden=TRUE}
#| label: hard coral
dat_hc <- dat |>
  filter(GROUP == "HC") |>
  droplevels()
```



# Fit models

::: {.panel-tabset}

## Binomial model

$$
\begin{align}
y_{i} &\sim{} Bin(\pi_{i}, n_{i})\\
log\left(\frac{\pi_i}{1-\pi_i}\right) &= \beta_0 + \beta_{i}\mathbf{X}\\
\beta_0 \sim{} N(0, 1)\\
\beta_{1-3} \sim{} N(0, 1)\\
\end{align}
$$

:::: {.panel-tabset}

### Define priors

```{r}
priors <- prior(normal(0,1), class = "Intercept")+
  prior(normal(0,3), class = "b")+ #fixed prior
  prior(student_t(3, 0, 1), class = "sd") #random prior

form <- bf(COUNT | trials(TOTAL)~ SAV_UPO + (1|site_name)+ (1|transect_id),
           family = binomial(link = "logit"))
```

### Fit prior only model

```{r}
#| label: Model1
#| cache: true
model1<- brm(form,
             data = dat_hc,
             prior = priors,
             sample_prior = "only",
             iter = 5000,
             warmup = 1000,
             chains = 3,
             cores = 3,
             thin = 5,
             refresh = 1000,
             control = list(adapt_delta = 0.99, max_treedepth = 20),
             backend = "rstan")
```
```{r}
#| fig.width: 10
#| fig.height: 5

model1 |>
  conditional_effects() |>
  plot( ask = FALSE) |>
  wrap_plots()
```


### Fit full model
```{r}
#| label: Model1_2
#| cache: true
model1 <- update(model1, sample_prior = "yes")
```






### MCMC sampling diagnostics


### Posterior probability checks
```{r}
model1 |>  SUYR_prior_and_posterior()
model1$fit |> stan_trace()
model1$fit |> stan_ac()
model1$fit |> stan_rhat()
model1$fit |> stan_ess()
```

```{r}
#| fig.width: 12
#| fig.height: 6
model1 |> pp_check(type = "dens_overlay", ndraws = 100)
resids <- model1 |> make_brms_dharma_res(integerResponse = FALSE)
wrap_elements(~testUniformity(resids))+
  wrap_elements(~plotResiduals(resids))+
  #wrap_elements(~plotResiduals(resids))+
  wrap_elements(~testDispersion(resids))
```

### Model validation



::::

:::

# Model posterior summaries

```{r}
model1 |> as_draws_df() |>
  summarise_draws(median, HDInterval::hdi,rhat,length,ess_bulk, ess_tail) |>
  knitr::kable() |> 
  head()
```

```{r}
plogis(-1.64) 
```




# Further explorations

::: {.panels-tabset}

## Pairwise constrasts
```{r}
model1 |> 
emmeans(~ SAV_UPO) |>
  regrid() |>
  pairs() |>
  gather_emmeans_draws() |>
  summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)) |>
 knitr::kable()
```
```{r}
model1 |> 
emmeans(~ SAV_UPO) |>
  regrid() |>
  regrid(trans = "log") |>
  pairs() |>
  gather_emmeans_draws() |>
  mutate(.value = exp(.value)) |> 
  summarise(median_hdci(.value),
            Pl = mean(.value < 1),
            Pg = mean(.value > 1)) |>
 knitr::kable()
```

## R2
```{r}
model1 |> bayes_R2(summary = FALSE, re_form = NA) |># ~(1|Reef_id) + (1| site_name)+ (1|transect_id)) |>
  median_hdci()|>
 knitr::kable() 
```
```{r}
model1 |> bayes_R2(summary = FALSE, re_form = ~(1|site_name) + (1| transect_id)) |> 
 median_hdci() |>
 knitr::kable()
```

```{r}
model1 |> bayes_R2(summary = FALSE, re_form = ~ (1|transect_id)) |>
  median_hdci()|>
 knitr::kable()
```


:::
# Summary figures
```{r}
#| fig.width: 10
#| fig.height: 5

model1 |>
  conditional_effects() |>
  plot( ask = FALSE) |>
  wrap_plots()
```

```{r}
g1<-
  model1 |>
  emmeans(~SAV_UPO, type = "response") |>
  as.data.frame() |>
  #gather_emmeans_draws() |>
  #mutate(fit = plogis(.value)) |>
  #summarise(median_hdci(fit)) |>
  ggplot(aes(y = prob, x = SAV_UPO))+
  geom_pointrange(aes(ymin = lower.HPD, ymax = upper.HPD))+
  scale_y_continuous("Coral Cover(%)", labels = scales::label_number(scale = 100))+
  #scale_x_discrete("COTs Outbreak", breaks = c(0,1), labels = c("No COTs", "COTS"))+
  theme_classic()
#| fig.width: 8
#| fig.height: 4
(g1 +annotate_npc("  a)", x = 0, y = 1, hjust = 0, vjust = 1))
```
```{r}
g2<-
   model1|>
  emmeans(~SAV_UPO,type = "response") |>
  regrid () |>
  pairs() |> #percentage changes to produce the plot 
  gather_emmeans_draws() |> #give me all 2400 of the values
  mutate(contrast = str_replace_all(contrast, "SAV_UPO", "")) |> #replace fYear with nothing 
  ggplot(aes(x = .value, y = contrast))+ #the column .value on x axis
  stat_halfeye(aes(fill = after_stat(level)), .width = c(0.66, 0.95,1))+ #produce density plots and fill it according to different proportion of the data
 # scale_fill_viridis_c() +
  scale_fill_brewer(palette = "Blues")+ #overwritten to have a different color pallet, color brewer is used for mapping you can change it to any color 
  geom_vline(xintercept = 0, linetype = "dashed")+ #dashed line 
  scale_x_continuous("Effect size")+ #providing nicer labels
  scale_y_discrete("") +
  theme_classic()
```
```{r}
#| fig.width: 12
#| fig.height: 8
(g1 +annotate_npc(" a)", x = 0, y = 1, hjust = 0, vjust = 1))+
(g2 + annotate_npc(" b)", x = 0, y = 1, hjust = 0, vjust = 1))
```
```{r}
sessionInfo()
```

